An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-01-25 Paper/response/response.tex

\documentclass[11pt,DIV=16]{scrartcl}

\usepackage[shortlabels]{enumitem}
\usepackage{xcolor}
\begin{document}

\section*{Reviewer 1}
Thanks for your careful evaluation and your comments and suggestions. Please find below answers to your remarks.
\begin{enumerate}[1)]
\item
\textit{page 3, equation (7). The authors should provide the definition of the matrix $\tilde{A}^\ell$.}\\
\textbf{The definition is now provided by the unnumbered equation between equation (5) and (6).}
\item
\textit{page 3, second column. Based on the definitions of $\mathbf{X}$ and $\mathbf{Y}$ given in the first column as well as equations (4) and (5), it is unclear how equation (6) could allow to forecast the signal at time $\frac{N-1+\ell}{f_s}$. Please give more details.}\\
\textbf{The paragraph has been rephrased, in order to make the forecasting procedure clearer. In particular, we detail how we obtain the estimation of $\mathbf{x}_{k+\ell}$ by recursively applying the linear relation (4). The estimation of the signal at time $\frac{N-1+\ell}{f_s}$ is then directly deduced from the last element of this vector.}
\item
\textit{page 7, second column. The authors wrote that the choice of the $M$ parameter is especially crucial when the the deterministic component of the signal is no longer stationary, like in the AHM model.
However, when they present an application of the proposed method to a signal satisfying the AHM model, neither they discuss how they select the $M$ value, nor they provide the analysis of the dependency of the performance of the proposed algorithm on the M value. Please include in section IV-C2 the evolution of the performance of the proposed method as a function of M and an explanation on how to tune its value in applications.}\\
\textbf{We have added the new paragraph IV-C2a where we discuss the influence of the parameter $M$ on the forecasting performance of an adaptive harmonic signal. In particular we mention that that choosing a too small value for $M$ does not provide enough information to forecast the signal satisfactorily, while a value of $M$ that is too large makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a
dozen oscillations is generally appropriate.}
\item
\textit{page 8, Table I. The authors present the performance of the extension method when applied to a toy AHM model. But they do not provide an insight of the impact, in this specific case, on the associated "boundary-free" TFR. Please include in this table the "performance index $D$" estimation proposed in section "Evaluation Metrics".}\\
\textbf{Table III now includes a new column containing the performance index $D$ of the boundary-free STFT associated with each of the extension method that we tested. For the sake of clarity, the performance of the boundary-free versions of the other TF representations (SST, RS and ConceFT) are not provided in this section. Their study is referred to the analysis of the real physiological signals (section IV-D).}
\item
\textit{page 10, Table III. Please provide the performance results associated with the first metric you proposed to use, the MSE. }\\
\textbf{This results have been added to Table III.}
\item
\textit{All real life signals proposed in the work are nonstationary signals which contain one single nonstationary component plus noise. In the toy example of a  AHM signal, section IV-C2, you presented the case of a signal containing two nonstationary components. Please add the example of a real life signal containing two or more simple nonstationary components plus noise.}\\
\textbf{Section IV-D1 now includes the application of \textsf{BoundEffRed} to a respiratory signal that contains not only the respiratory cycle, but also a
cardiogenic artifact. We then show the ability of our algorithm to work on
signals containing several nonstationary components. Another application to an atrial blood pressure signal is provided in section III of the Supplementary Materials. This signal also contains both simultaneously a cardiac component and a respiratory component.}
\end{enumerate}

### 2021-01-25 Paper/BoundEffRed.tex
paper

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR). There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, the low computational time allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. An additional application to ECG is included in the Supplementary Materials.

>>>>

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
% 
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}. 
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, the low computational time allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. Additional applications to ECG and atrial blood pressure signals are included in the Supplementary Materials.

### 2021-01-25 Paper/results.tex
paper and results
paper

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$. 
%
{\color{red}
In practice, the asymptotic behavior of {\sf SigExt}, illustrated here on a synthetic signal, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K\gg L$.
}

@@

{\color{red}
\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to four different values of the subsignals length $M$. Note that choosing a too small value for $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a value of $M$ that is too large makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.
}

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & $0.0091 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & $0.0056 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M\!=\!1500$)} & $0.634$ & $0.907 \pm  0.377$ & $0.0065 \pm  0.0023$\\
   \hline
   {\sf SigExt ($M\!=\!3000$)} & $2.881$ & $5.365 \pm  3.919$ & $0.0117 \pm  0.0038$\\
   \hline
   Symmetric &  $<0.001$ & $5.452 \pm  0.002$ & $1.8209 \pm  0.0025$\\
   \hline
   EDMD & $1.587$ & $0.983  \pm  0.004$ & $0.1145 \pm  0.0018$\\
   \hline
   GPR & $515.787$ & $0.479  \pm  0.166$ & $0.0119 \pm  0.0037$\\
   \hline
   TBATS  & {\color{blue}$x$} & {\color{blue}$x \pm  x$} & {\color{blue}$x \pm  x$}\\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

{\color{red}
Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also the cardiogenic artifact which reflects the hemodynamic component. Fig.~\ref{fig:resp.2} displays the boundary-free SST of a 180-second-long extract of this signal, sample at $125$~Hz, obtained via {\sf BoundEffRed}. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. The performance index of this representation takes the value $D=0.314$. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $69\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\caption{Boundary-free SST of a two-component respiratory signal.}
\label{fig:resp.2}
\end{figure} 
}

>>>>

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$. 
%
In practice, the asymptotic behavior of {\sf SigExt}, illustrated here on a synthetic signal, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K\gg L$.

@@

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to four different values of the subsignals length $M$. Note that choosing a too small value for $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a value of $M$ that is too large makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results. Indeed, when the subsignal length $M$ is optimally chosen, {\sf SigExt} performance is of the same order as that of ore sophisticated methods, like GPR or TBATS. Besides, even though these methods could be slightly more robust to perturbations, they are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & $0.0091 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & $0.0056 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M\!=\!1500$)} & $0.634$ & $0.907 \pm  0.377$ & $0.0065 \pm  0.0023$\\
   \hline
   {\sf SigExt ($M\!=\!3000$)} & $2.881$ & $5.365 \pm  3.919$ & $0.0117 \pm  0.0038$\\
   \hline
   Symmetric &  $<0.001$ & $5.452 \pm  0.002$ & $1.8209 \pm  0.0025$\\
   \hline
   EDMD & $1.587$ & $0.983  \pm  0.004$ & $0.1145 \pm  0.0018$\\
   \hline
   GPR & $515.787$ & $0.479  \pm  0.166$ & $0.0119 \pm  0.0037$\\
   \hline
   TBATS  & $1776.778$ & $0.686 \pm  0.467$ & $0.3847 \pm  0.1397$\\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also the cardiogenic artifact which reflects the hemodynamic component. Fig.~\ref{fig:resp.2} displays the boundary-free SST of a 180-second-long extract of this signal, sample at $125$~Hz, obtained via {\sf BoundEffRed}. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. The performance index of this representation takes the value $D=0.314$. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $69\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\caption{Boundary-free SST of a two-component respiratory signal.}
\label{fig:resp.2}
\end{figure}
